### 2020-06-22 submission/manuscript.Rmd
Edits after updating figures, Nick's review, and my review of manuscript

## Abstract
The microbiota plays a key role in determining susceptibility to *Clostridioides difficile* infections (CDIs). However, much of the mechanistic work examining CDIs in mouse models use a single university colony or vendor, which have lower inter-individual microbiota variation compared to humans. We treated mice from 6 different colony sources (2 University of Michigan colonies and 4 vendors) with a single clindamycin dose, followed by *C. difficile* challenge 1 day later and measured *C. difficile* colonization levels through 9 days post-infection. The microbiota was profiled throughout the experiment via 16S rRNA gene sequencing analysis to examine variation across colony sources and alterations due to clindamycin treatment and *C. difficile* challenge. While all sources of mice were colonized 1-day post-infection, variation in *C. difficile* colonization levels emerged from days 3-7 post-infection with 3 sources colonized with *C. difficile* for longer and at higher levels. We identified bacterial taxa with different relative abundances across colony sources throughout the experiment, as well as taxa that were consistently impacted by clindamycin treatment. We created bacterial community-based logistic regression models that successfully classified mice based on their day 7 *C. difficile* colonization status. Importantly, after examining the taxa that were most important to the classification models, we identified a subset of key taxa that varied across colony sources (*Bacteroides*, *Deferribacteraceae*), were altered by clindamycin (*Porphyromonadaceae*, *Ruminococcaceae*), or both (*Enterobacteriaceae*, *Enterococcus*, *Bifidobacteriaceae*, *Coriobacteriaceae*, *Lachnospiraceae*, and *Verrucomicrobiaceae*). These results suggest the response of the initial gut microbiota to clindamycin treatment influences *C. difficile* 630 colonization dynamics.

@@

Mouse models of CDIs have been a great tool for understanding *C. difficile* pathogenesis [@Hutton2014]. The number of CDI mouse model studies has grown substantially since Chen et al. published their C57BL/6 model in 2008, which disrupted the gut microbiota with antibiotics to enable *C. difficile* colonization, diarrhea, and weight loss [@Chen2008]. CDI mouse models have been used to examine translationally relevant questions regarding *C. difficile*, including the role of the microbiota and efficacy of potential therapeutics for treating CDIs [@Best2012]. However, microbiome variation between lab mice is much less than the variation observed between humans [@Baxter2014; @Nagpal2018]. Additionally, studying the contribution of the microbiota to a particular disease phenotype in one set of lab mice after the same perturbation could yield a number of findings of which only a fraction may be driving the phenotype.

@@

```{r cfu}
cfu_stats_all <- read_tsv("../data/process/cfu_stats_all_days.tsv")
padj_d1_cfu <- cfu_stats_all %>% filter(day == 1) %>% pull(p.value.adj) 
mean_d1 <- cfu_stats_all %>% filter(day == 1) %>% select(-day, -statistic, -p.value, -parameter, -method, -p.value.adj) %>% pivot_longer(everything(), values_to = "mean_cfu", names_to = "Source") %>% arrange(mean_cfu)
min_mean_d1 <- mean_d1 %>%  select(mean_cfu) %>% min
max_mean_d1 <- mean_d1 %>% select(mean_cfu) %>% max

cfu_stats_sig <- cfu_stats_all %>% filter(p.value.adj < 0.05)
max_padj_d3to7 <- cfu_stats_sig %>% select(p.value.adj) %>% max
```
## Results

**Clindamycin treatment renders all mice susceptible to *C. difficile* 630 colonization regardless of colony source.** To test how the microbiotas of mice from different colony sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan (the Young and Schloss lab colonies), the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan). These 4 vendors were chosen because they represent commonly used vendors for CDI studies in mice [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019]. After a 13-day acclimation period for the mice ordered from vendors, all mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C. difficile* 630 spores (Fig. 1A). Clindamycin was chosen because we have previously demonstrated mice are rendered susceptible, but consistently cleared the CDI within 9 days [@Jenior2018; @Tomkovich2019], clindamycin is frequently implicated with human CDIs [@Guh2018], and is also part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008]. The day after infection, *C. difficile* was detectable in all mice at a similar level (Mean CFU range: `r format(min_mean_d1, digits = 2, scientific=TRUE)`-`r format(max_mean_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of colony source (Fig. 1B). Interestingly, variation in *C. difficile* CFU levels across sources of mice emerged from days 3-7 post-infection (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig. 1B and Table S1), suggesting mouse colony source is associated with *C. difficile* clearance. We conducted two experiments approximately 3 months apart and combined the data because the colonization dynamics across sources of mice varied similarily in both experiments (Fig. S1). Although *C. difficile* 630 causes mild symptoms in mice comparied to other *C. difficile* strains [@Theriot2011], we also saw that weight change significantly varied across sources of mice with the most weight lost two days post-infection (Fig. 1C and Table S2). Interestingly, mice ordered from Jackson, Taconic, and Envigo tended to lose the most weight, have higher *C. difficile* CFU levels and take longer to clear the infection compared to the other sources of mice, which was particularly evident 7 days post-infection (Fig. 1C and D). Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection. Thus, clindamycin rendered all mice susceptible to *C. difficile* 630 colonization, regardless of colony source, but variation across sources emerged with 3 out of 6 sources taking longer to clear *C. difficile*.

@@

shannon_stats_all <- read_tsv("../data/process/shannon_stats_all_days.tsv")
padj_dn1_shannon <- shannon_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
```
**Bacterial communities vary across mouse colony sources.** Given the well known variation in mouse microbiomes across vendors and university colonies [@Alegre2019], we hypothesized that the variation in *C. difficile* clearance could be explained by microbiota variation across the 6 sources. We used 16S rRNA gene sequencing to characterize the fecal bacterial communities from the mice over the course of the experiment. Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity measures across the 6 sources of mice. Examining the bacterial communities at baseline, prior to clindamycin treatment there was a significant difference in the number of observed OTUs (*P*~FDR~ = `r padj_dn1_richness`), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_shannon`) across sources of mice (Fig. 2A-B and Table S3). As expected, clindamycin treatment decreased richness and Shannon diversity across all sources of mice, and communities started to recover 1 day post-infection (Fig. 2C-D). Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after both clindamycin and *C. difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other groups (Fig 2C-F and Table S4). While Charles River mice had more diverse microbiotas, Young and Schloss lab mice were also able to clear *C. difficile* faster, suggesting microbiota diversity alone does not explain the observed variation in *C. difficile* colonization across vendors.
```{r adonis}
adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
r_sq_ordered_adonis <- adonis_stats_all %>% arrange(desc(r_sq))
rsq_adonis_source <- r_sq_ordered_adonis %>% pull(r_sq) %>% max
p_adonis_source <- r_sq_ordered_adonis %>% pull(p) %>% max

@@

source_adonis_dn1 <- read_tsv("../data/process/adonis_dn1_source.tsv")
adonis_schloss_rsq_sum <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(r_sq) %>% sum
p_adonis_schloss_max <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(p) %>% max
adonis_young_rsq_sum <- source_adonis_dn1 %>% filter(source == "Young") %>% select(r_sq) %>% sum
p_adonis_young_max <- source_adonis_dn1 %>% filter(source == "Young") %>% select(p) %>% max
```
Next, we compared the bacterial communities from the 6 colonies over the course of the experiment using principal coordinate analysis (PCoA) of the Theta YC distances. Permutational multivariate analysis of variance (PERMANOVA) analysis revealed colony source was the major factor explaining the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment (Movie S1 and Table S5). Since, the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the bacterial communities at baseline (day -1), after clindamycin treatment (day 0), and post-infection (day 1). For all 3 timepoints, source and the interaction with cage significantly explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, `r d0_rsq_sum`, `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig. 3 and Table S6). We also compared baseline communities across the 2 experiments, and found experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig. S2 and Table S7), suggesting vendor communities were relatively stable between the 2 experiments. Thus, mouse colony source was the factor that explained the most variation observed in the bacterial communities. Importantly with the exception of the 2 University colonies, the community of each source clustered apart from one another suggesting each community had a unique response to clindamycin treatment and *C. difficile* challenge.
```{r source_taxa}
#Number of different otus and families across sources at day -1, 0, and 1 timepoints
source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
source_families <- read_tsv("../data/process/family_stats_dn1to1_combined.tsv")
source_families_dn1 <- source_families %>% filter(day == -1) %>% nrow()
source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
source_families_d0 <- source_families %>% filter(day == 0) %>% nrow()
source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()
source_families_d1 <- source_families %>% filter(day == 1) %>% nrow()

@@

#Families shared across baseline, post-clindamycin, and post-infection timepoints
source_f_dn1 <- source_families %>% filter(day == -1) %>% pull(family)
source_f_d0 <- source_families %>% filter(day == 0) %>% pull(family)
source_f_d1 <- source_families %>% filter(day == 1) %>% pull(family)
shared_source_families_dn1to1 <- length(intersect_all(source_f_dn1, source_f_d0, source_f_d1))
```
After finding differences at the community level, we next identified the taxa that varied across sources of mice over the initial days of the experiment. We examined bacterial relative abundances at the operational taxonomic unit (OTU) and family levels, expecting the number of differences to be reduced at the family level due to the nature of bacterial taxonomy [@Nguyen2015]. Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs and only `r source_families_dn1` families (Table S8-9) with relative abundances that varied across colony sources. Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs and `r source_families_d0` families (Table S8-9). After *C. difficile* challenge, there were `r source_otus_d1` OTUs and `r source_families_d1` families (Table S8-9) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment. In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 4A-C) and `r shared_source_families_dn1to1` families with relative abundances that consistently varied across colony sources (Fig. 4D-F). Importantly, some of the taxa that consistently varied across sources also shifted with clindamycin treatment. For example, *Proteus* increased after clindamycin treatment, but only in Taconic mice. *Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment. In summary, mouse bacterial communities significantly varied according to colony source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C. difficile* challenge.

@@

clindamycin_families <- read_tsv("../data/process/family_stats_dn1to0.tsv")
clindamycin_families_sig <- clindamycin_families %>% filter(p.value.adj < 0.05) %>% nrow()
```
**Clindamycin treatment alters a subset of taxa that were found in all colony sources.** Although there were bacteria that consistently varied across colony sources, we also wanted to identify the bacteria that shifted after clindamycin treatment, regardless of colony source. By analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment, we identified `r clindamycin_otus_sig` OTUs and `r clindamycin_families_sig` families that were altered after clindamycin treatment (Fig. 5 and Table S10-11). Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that consistently varied across groups over the initial 3 timepoints of our experiment, we found 3 OTUs (*Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) and 3 families (*Porphyromonadaceae*, *Enterococcaceae*, *Lachnospiraceae*) overlapped (Fig. 4, Fig. 5C-D). These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all colony sources and only a subset of the taxa also varied across colony sources.

@@

By examining the *C. difficile* colonization dynamics within mice from 6 different colony sources after perturbing the microbiota with clindamycin treatment, we were able to identify bacterial taxa that were unique to sources throughout the experiment as well as taxa that were universally impacted by clindamycin. We built L2 logistic regression models with baseline, post-clindamycin treatment, and post-infection fecal community data that successfully predicted *C. difficile* colonization status 7 days after infection better than random chance. We identified *`r list_key_otus`*, *`r list_key_families`* (Fig. 8, Fig. S7) as candidate bacteria within these communities that were influencing variation in *C. difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by colony source, were impacted by clindamycin treatment, or both. Overall, our results demonstrate clindamycin is sufficient to render mice from multiple sources susceptible to CDI and only a subset of the interindividual microbiota variation across mice from different sources was associated with the time needed to clear *C. difficile*.

>>>>

## Abstract
The microbiota plays a key role in determining susceptibility to *Clostridioides difficile* infections (CDIs). However, much of the mechanistic work examining CDIs in mouse models use a single university colony or vendor. We treated mice from 6 different colony sources (2 University of Michigan colonies and 4 vendors) with a single clindamycin dose, followed by *C. difficile* challenge 1 day later and measured *C. difficile* colonization levels through 9 days post-infection. The microbiota was profiled via 16S rRNA gene sequencing analysis to examine variation across colony sources and alterations due to clindamycin treatment and *C. difficile* challenge. While all sources of mice were colonized 1-day post-infection, variation in *C. difficile* colonization levels emerged from days 3-7 post-infection with 3 sources colonized with *C. difficile* for slightly longer and at higher levels. We identified bacterial taxa with different relative abundances across colony sources throughout the experiment, as well as taxa that were consistently impacted by clindamycin treatment in all sources of mice. We created logistic regression models that successfully classified mice based on whether they cleared *C. difficile* by 7 days post-infection using baseline, post-clindamycin, and post-infection community composition data. After examining the taxa that were most important to the classification models, we identified a subset of key taxa that varied across colony sources (*Bacteroides*, *Deferribacteraceae*), were altered by clindamycin (*Porphyromonadaceae*, *Ruminococcaceae*), or both (*Enterobacteriaceae*, *Enterococcus*, *Bifidobacteriaceae*, *Coriobacteriaceae*, *Lachnospiraceae*, and *Verrucomicrobiaceae*). These results suggest the response of the initial gut microbiota to clindamycin treatment influences *C. difficile* 630 colonization dynamics.

@@

Mouse models of CDIs have been a great tool for understanding *C. difficile* pathogenesis [@Hutton2014]. The number of CDI mouse model studies has grown substantially since Chen et al. published their C57BL/6 model in 2008, which disrupted the gut microbiota with antibiotics to enable *C. difficile* colonization and symptoms such as diarrhea and weight loss [@Chen2008]. CDI mouse models have been used to examine translationally relevant questions regarding *C. difficile*, including the role of the microbiota and efficacy of potential therapeutics for treating CDIs [@Best2012]. However, microbiome variation between lab mice is much less than the variation observed between humans [@Baxter2014; @Nagpal2018]. Additionally, studying the contribution of the microbiota to a particular disease phenotype in one set of lab mice after the same perturbation could yield a number of findings of which only a fraction may be driving the phenotype.

@@

```{r cfu}
cfu_stats_all <- read_tsv("../data/process/cfu_stats_all_days.tsv")
padj_d1_cfu <- cfu_stats_all %>% filter(day == 1) %>% pull(p.value.adj) 
median_d1 <- cfu_stats_all %>% filter(day == 1) %>% select(-day, -statistic, -p.value, -parameter, -method, -p.value.adj) %>% pivot_longer(everything(), values_to = "median_cfu", names_to = "Source") %>% arrange(median_cfu)
min_median_d1 <- median_d1 %>%  select(median_cfu) %>% min
max_median_d1 <- median_d1 %>% select(median_cfu) %>% max

cfu_stats_sig <- cfu_stats_all %>% filter(p.value.adj < 0.05)
max_padj_d3to7 <- cfu_stats_sig %>% select(p.value.adj) %>% max

#Percent of mice colonized with C. difficile each day
metadata <- read_tsv("../data/process/processed_metadata.tsv")
#Add clearance_status_dx columns for the rest of the days in addition to d7
metadata_cfu_status <- metadata %>% 
  mutate(clearance_status_d7 = case_when(cfu_d7 > 0 ~ "colonized",
                                         cfu_d7 == 0 ~ "cleared", #changed to match plot labeling used above
                                         cfu_d7 == NA ~ "NA"),
         clearance_status_d1 = case_when(cfu_d1 > 0 ~ "colonized",
                                         cfu_d1 == 0 ~ "cleared", 
                                         cfu_d1 == NA ~ "NA"),
         clearance_status_d2 = case_when(cfu_d2 > 0 ~ "colonized",
                                         cfu_d2 == 0 ~ "cleared", 
                                         cfu_d2 == NA ~ "NA"),
         clearance_status_d3 = case_when(cfu_d3 > 0 ~ "colonized",
                                         cfu_d3 == 0 ~ "cleared", 
                                         cfu_d3 == NA ~ "NA"),
         clearance_status_d4 = case_when(cfu_d4 > 0 ~ "colonized",
                                         cfu_d4 == 0 ~ "cleared", 
                                         cfu_d4 == NA ~ "NA"),
         clearance_status_d5 = case_when(cfu_d5 > 0 ~ "colonized",
                                         cfu_d5 == 0 ~ "cleared", 
                                         cfu_d5 == NA ~ "NA"),
         clearance_status_d6 = case_when(cfu_d6 > 0 ~ "colonized",
                                         cfu_d6 == 0 ~ "cleared", 
                                         cfu_d6 == NA ~ "NA"),
         clearance_status_d8 = case_when(cfu_d8 > 0 ~ "colonized",
                                         cfu_d8 == 0 ~ "cleared", 
                                         cfu_d8 == NA ~ "NA"),
         clearance_status_d9 = case_when(cfu_d9 > 0 ~ "colonized",
                                         cfu_d9 == 0 ~ "cleared", 
                                         cfu_d9 == NA ~ "NA"))

#Percent mice colonized each day, 2 experiments combined----
count_colonized <- function(timepoint, clearance_status_dx){
  metadata_cfu_status %>% 
    filter(day == timepoint, !is.na(cfu)) %>% #get rid of NA values
    count({{ clearance_status_dx }}) %>% 
    mutate(day = timepoint) %>% 
    rename(cfu_status = {{ clearance_status_dx }})
}

#Figure out number of mice that either cleared C. difficile or were still colonized with C. difficile on each day
day_1 <- count_colonized(1, clearance_status_d1)
day_2 <- count_colonized(2, clearance_status_d2)
day_3 <- count_colonized(3, clearance_status_d3)
day_4 <- count_colonized(4, clearance_status_d4)
day_5 <- count_colonized(5, clearance_status_d5)
day_6 <- count_colonized(6, clearance_status_d6)
day_7 <- count_colonized(7, clearance_status_d7)
day_8 <- count_colonized(8, clearance_status_d8)
day_9 <- count_colonized(9, clearance_status_d9)

#Combine into a data frame and figure out percent colonized per day
percent_colonized_day <- rbind(day_1, day_2, day_3, day_4, day_5, day_6, day_7, day_8, day_9) %>% 
  group_by(day) %>% 
  mutate(total_n = sum(n)) %>% #Figure out total n per day (sum of cleared and colonized counts)
  filter(cfu_status == "colonized") %>% #Only need colonized rows to figure out percent colonized
  mutate(percent_colonized = (n/total_n) * 100)#Add new column for percent_colonized
percent_colonized_d7 <- percent_colonized_day %>% filter(day == 7) %>% 
  pull(percent_colonized)
n_day1 <- percent_colonized_day %>% filter(day == 1) %>% 
  pull(total_n)
n_day2 <- percent_colonized_day %>% filter(day == 2) %>% 
  pull(total_n)
n_day3 <- percent_colonized_day %>% filter(day == 3) %>% 
  pull(total_n)
n_day4 <- percent_colonized_day %>% filter(day == 4) %>% 
  pull(total_n)
n_day5 <- percent_colonized_day %>% filter(day == 5) %>% 
  pull(total_n)
n_day6 <- percent_colonized_day %>% filter(day == 6) %>% 
  pull(total_n)
n_day7 <- percent_colonized_day %>% filter(day == 7) %>% 
  pull(total_n)
n_day8 <- percent_colonized_day %>% filter(day == 8) %>% 
  pull(total_n)
n_day9 <- percent_colonized_day %>% filter(day == 9) %>% 
  pull(total_n)
```
## Results

**Clindamycin treatment renders all mice susceptible to *C. difficile* 630 colonization regardless of colony source.** To test how the microbiotas of mice from different colony sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan (the Young and Schloss lab colonies), the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan). These 4 vendors were chosen because they represent commonly used vendors for CDI studies in mice [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019]. After a 13-day acclimation period for the mice ordered from vendors, all mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C. difficile* 630 spores (Fig. 1A). Clindamycin was chosen because we have previously demonstrated mice are rendered susceptible, but consistently cleared the CDI within 9 days [@Jenior2018; @Tomkovich2019], clindamycin is frequently implicated with human CDIs [@Guh2018], and is also part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008]. The day after infection, *C. difficile* was detectable in all mice at a similar level (median CFU range: `r format(min_median_d1, digits = 2, scientific=TRUE)`-`r format(max_median_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of colony source (Fig. 1B). Interestingly, variation in *C. difficile* CFU levels across sources of mice emerged from days 3-7 post-infection (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig. 1B and Table S1), suggesting mouse colony source is associated with *C. difficile* clearance. We conducted two experiments approximately 3 months apart and while the colonization dynamics were similar across most sources of mice, there was some variation between the 2 experiments, particularly for the Schloss and Envigo mice (Fig. S1A-B, Fig. 1D). Although *C. difficile* 630 causes mild symptoms in mice comparied to other *C. difficile* strains [@Theriot2011], we also saw that weight change significantly varied across sources of mice with the most weight lost two days post-infection (Fig. 1C, E and Table S2). Interestingly, mice ordered from Jackson, Taconic, and Envigo tended to lose more weight (although there was variation between experiments with Schloss and Envigo mice), have higher *C. difficile* CFU levels and take longer to clear the infection compared to the other sources of mice, which was particularly evident 7 days post-infection (Fig. 1B-E), when `r percent_colonized_d7` of the mice were still colonized with *C. difficile* (Fig. S1C). By 9 days post-infection the majority of the mice from all sources had cleared *C. difficile* (Fig. S1C) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment. Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment. Thus, clindamycin rendered all mice susceptible to *C. difficile* 630 colonization, regardless of colony source, but variation across sources emerged with 3 out of 6 sources taking longer to clear *C. difficile*.

@@

shannon_stats_all <- read_tsv("../data/process/shannon_stats_all_days.tsv")
padj_dn1_shannon <- shannon_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
```
**Bacterial communities consistently vary across mouse colony sources despite antibiotic and infection perturbations.** Given the well known variation in mouse microbiomes across vendors and university colonies [@Alegre2019], we hypothesized that the variation in *C. difficile* clearance could be explained by microbiota variation across the 6 sources. We used 16S rRNA gene sequencing to characterize the fecal bacterial communities from the mice over the course of the experiment. Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity measures across the 6 sources of mice. Examining the bacterial communities at baseline, prior to clindamycin treatment there was a significant difference in the number of observed OTUs (*P*~FDR~ = `r padj_dn1_richness`), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_shannon`) across sources of mice (Fig. 2A-B and Table S3). As expected, clindamycin treatment decreased richness and Shannon diversity across all sources of mice, and communities started to recover 1 day post-infection (Fig. 2C-D). Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after both clindamycin and *C. difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other groups (Fig 2C-F and Table S4). While Charles River mice had more diverse microbiotas, Young and Schloss lab mice were also able to clear *C. difficile* faster, suggesting microbiota diversity alone does not explain the observed variation in *C. difficile* colonization across vendors.
```{r adonis}
adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
r_sq_ordered_adonis <- adonis_stats_all %>% arrange(desc(r_sq))
rsq_adonis_source <- r_sq_ordered_adonis %>% pull(r_sq) %>% max
p_adonis_source <- r_sq_ordered_adonis %>% pull(p) %>% max

@@

source_adonis_dn1 <- read_tsv("../data/process/adonis_dn1_source.tsv")
adonis_schloss_rsq_sum <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(r_sq) %>% sum
p_adonis_schloss_max <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(p) %>% max
adonis_young_rsq_sum <- source_adonis_dn1 %>% filter(source == "Young") %>% select(r_sq) %>% sum
p_adonis_young_max <- source_adonis_dn1 %>% filter(source == "Young") %>% select(p) %>% max
```
Next, we compared the bacterial communities from the 6 colonies over the course of the experiment using principal coordinate analysis (PCoA) of the Theta YC distances. Permutational multivariate analysis of variance (PERMANOVA) analysis revealed colony source was the major factor explaining the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment (Movie S1 and Table S5). Since, the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the bacterial communities at baseline (day -1), after clindamycin treatment (day 0), and post-infection (day 1). For all 3 timepoints, source and the interaction with cage significantly explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, `r d0_rsq_sum`, `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig. 3 and Table S6). We also compared baseline communities across the 2 experiments, and found experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig. S2 and Table S7), although most of the vendors also clustered by experiment, suggesting there was some community variation between the 2 experiments within each vendor. Thus, mouse colony source was the factor that explained the most variation observed in the bacterial communities. Importantly with the exception of the 2 University colonies, the community of each source clustered apart from one another suggesting each community had a unique response to clindamycin treatment and *C. difficile* challenge.

Since there was some variation in microbiota communities between experiments at baseline, we next looked at how similar the communities were within the same source and between sources in response to clindamycin treatment and *C. difficile* challenge (Fig. 4) . The baseline communities varied most between experiments for Schloss, Young, and Envigo mice and variation between sources of mice was high (Fig. 4A). Clindamycin treatment reduced variation between experiments within Schloss, Young and Jackson mice and some of the variation between sources diminished (Fig. 4B). Post-infection, the community variation started to increase within sources of mice and variation between sources of mice started to return (Fig. 4C). By using mice from multiple sources we were able to increase the number of microbiota communities we tested with the clindamycin *C. difficile* colonization mouse model.
```{r source_taxa}
#Number of different otus and families across sources at day -1, 0, and 1 timepoints
source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
source_families <- read_tsv("../data/process/family_stats_dn1to1_combined.tsv")
source_families_dn1 <- source_families %>% filter(day == -1) %>% nrow()
source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
source_families_d0 <- source_families %>% filter(day == 0) %>% nrow()
source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()
source_families_d1 <- source_families %>% filter(day == 1) %>% nrow()

@@

#Families shared across baseline, post-clindamycin, and post-infection timepoints
source_f_dn1 <- source_families %>% filter(day == -1) %>% pull(family)
source_f_d0 <- source_families %>% filter(day == 0) %>% pull(family)
source_f_d1 <- source_families %>% filter(day == 1) %>% pull(family)
shared_source_families_dn1to1 <- length(intersect_all(source_f_dn1, source_f_d0, source_f_d1))
```
After finding differences at the community level, we next identified the taxa that varied across sources of mice over the initial days of the experiment. We examined bacterial relative abundances at the operational taxonomic unit (OTU) and family levels, expecting the number of differences to be reduced at the family level due to the nature of bacterial taxonomy [@Nguyen2015]. Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs and `r source_families_dn1` families (Table S8-9) with relative abundances that varied across colony sources. Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs and `r source_families_d0` families (Table S8-9). After *C. difficile* challenge, there were `r source_otus_d1` OTUs and `r source_families_d1` families (Table S8-9) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment. In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 5A-C) and `r shared_source_families_dn1to1` families with relative abundances that consistently varied across colony sources (Fig. 5D-F). Importantly, some of the taxa that consistently varied across sources also shifted with clindamycin treatment. For example, *Proteus* increased after clindamycin treatment, but only in Taconic mice. *Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment. In summary, mouse bacterial communities significantly varied according to colony source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C. difficile* challenge.

@@

clindamycin_families <- read_tsv("../data/process/family_stats_dn1to0.tsv")
clindamycin_families_sig <- clindamycin_families %>% filter(p.value.adj < 0.05) %>% nrow()
```
**Clindamycin treatment alters a subset of taxa that were found in all colony sources.** Although there were bacteria that consistently varied across colony sources, we also wanted to identify the bacteria that shifted after clindamycin treatment, regardless of colony source. By analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment, we identified `r clindamycin_otus_sig` OTUs and `r clindamycin_families_sig` families that were altered after clindamycin treatment (Fig. 6 and Table S10-11). Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that consistently varied across groups over the initial 3 timepoints of our experiment, we found 3 OTUs (*Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) and 3 families (*Porphyromonadaceae*, *Enterococcaceae*, *Lachnospiraceae*) overlapped (Fig. 5, Fig. 6C-D). These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all colony sources and only a subset of the taxa also varied across colony sources.

@@

By examining the *C. difficile* colonization dynamics within mice from 6 different colony sources after perturbing the microbiota with clindamycin treatment, we were able to identify bacterial taxa that were unique to sources throughout the experiment as well as taxa that were universally impacted by clindamycin. We built L2 logistic regression models with baseline, post-clindamycin treatment, and post-infection fecal community data that successfully predicted *C. difficile* colonization status 7 days after infection better than random chance. We identified *`r list_key_otus`*, *`r list_key_families`* (Fig. 8, Fig. S6) as candidate bacteria within these communities that were influencing variation in *C. difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by colony source, were impacted by clindamycin treatment, or both. Overall, our results demonstrate clindamycin is sufficient to render mice from multiple sources susceptible to CDI and only a subset of the interindividual microbiota variation across mice from different sources was associated with the time needed to clear *C. difficile*.

Other groups have taken similar approaches by using mice from multiple colony sources to identify bacteria that either promote colonization resistance or increase susceptibility to enteric infections [@Velazquez2019; @EtienneMesmin2017; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018]. For example, in the context of *Salmonella* infections, *Enterobacteriaceae* and segmented filamentous bacteria have emerged as protective [@Velazquez2019; @Lai2020]. A previous study with *C. difficile* identified an endogenous protective *C. difficile* strain LEM1 that bloomed after antibiotic treatment in mice from Jackson or Charles River Laboratories, but not Taconic that protected mice against the more toxigenic *C. difficile* VPI10463 [@EtienneMesmin2017]. Given that we ordered mice from the same vendors, we checked all mice for endogenous *C. difficile* by plating stool samples that were collected after clindamycin treatment. However, we did not identify any endogenous *C. difficile* strains prior to challenge, suggesting there were no endogenous protective strains in the mice we received and other bacterial taxa mediated the variation in *C. difficile* colonization across sources. Although all mice were susceptible to *C. difficile* colonization, by following colonization over time we found Jackson, Taconic, and Envigo mice remained colonized beyond 7 days post-infection. We identified a subset of bacteria that were important in predicting whether a mouse was still colonized with *C. difficile* 7 days post-infection. These results suggest a subset of the bacterial community is responsible for determining the length of time needed to clear *C. difficile* colonization.
